Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-06 paper/main.tex
pass through section 2

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}

@@

\jan{We also mention before that existing imperative tools don't allow
  for user interaction. We should pick this up again and say that we
  can do it. But then we have to mention the PLDI paper. My suggestion:
  \begin{itemize}
  \item mention PLDI'14 in the paragraph about the second category and
    say it has great flexibility for manual bound derivation but
    almost no support for automation
  \item just add another bullet to say that we have the same
    flexibility for manual bound derivation as in PLDI'14 and that
    manual and automatic bounds and be composed
  \end{itemize}
}

@@

In this section, we informally introduce the automatic amortized
analysis (AAA) for Clight programs.  Some example bounds are derived
step by step following our algorithm.

\subsection{Bounding resource usage with potential functions}

@@

The reasoning presented above is very general and flexible.  But in
practice, to make the reasoning tractable on large code bases, we would
like to infer automatically the potential functions.  The first step
towards this goal is to reduce our search space by fixing the shape
of potential functions $\Phi$.  We settle on a form that allows
\emph{linear programming} (LP) to help inference:
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom(\state)}
  q_{(x,y)} \cdot |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  We assumed
that the program state is a simple map from program variables to integers
and we treat constants as program variables (that cannot be changed
by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
the sum.  We then develop an inference rule for each syntactic construct
that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
the previous section.

@@

\subsection{Compositionality in practice}

With two concrete examples from open-source projects we will try to
show the reader that compositionality is indeed useful in the analysis
of common code.

The following example is typical in cryptographic primitives
implementations.  These primitives work by block, so when data of
arbitrary length needs to be processed, it is munched in blocks and
the leftovers are stored in a buffer for future use when more data
is available.  This looping pattern is particularly well handled by
our tool because the LP solver is leveraged to generate a tight bound.
The optimization problem it is given encodes that constant bounds are
preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
AAA is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
+ \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
counted as constant.
%
\begin{lstlisting}
for (; l >= 8; l -= 8)
    // process one block
    tick(N);
for (; l > 0; l--)
    // save leftovers
    tick(1);
\end{lstlisting}
%
It is worth taking some time to explain the constant contribution
to the resource bound when $N<8$.  Remark first that the cost of
the second loop is $|[0,l]|$, and after the first one, we still have
$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
potential units instead.  This explains the somewhat intriguing value
of the constant term in the bound.  All this reasoning is done within
our tool automatically and simply relies on general rules presented
in further sections.

@@

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && nondet());
    do { h--; tick(1); } while (h > l && nondet());
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  However, all available
tools for C will derive a quadratic bound instead.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  But on the other hand, current
techniques will try to find one linear ranking-function for each loop
and combine them multiplicatively or additively, resulting in a loose
quadratic bound.

>>>>

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\jan{We can say that our technique is the only one that reduces the problem to
  LP solving. I used that as a selling point in the DARPA-STAC proposal.}

@@

\jan{We also mention before that existing imperative tools don't allow
  for user interaction. We should pick this up again and say that we
  can do it. But then we have to mention the PLDI paper. My suggestion:
  \begin{itemize}
  \item mention PLDI'14 in the paragraph about the second category and
    say it has great flexibility for manual bound derivation but
    almost no support for automation
  \item just add another bullet to say that we have the same
    flexibility for manual bound derivation as in PLDI'14 and that
    manual and automatic bounds can be composed
  \end{itemize}
}

@@

In this section, we informally introduce the automatic amortized
analysis (AAA) for Clight programs and demonstrate the technique with
example bounds that we derive step-by-step following our analysis
system.

\jan{Watch out with the headings. Either upper case or lower case, but
  don't mix.}

\subsection{Bounding Resource Usage with Potential Functions}

@@

The reasoning presented above is very general and flexible.  But in
practice, to make the reasoning tractable on large code bases, we would
like to automatically infer the potential functions.  The first step
towards this goal is to reduce our search space by fixing the shape
of potential functions $\Phi$.  We settle on a form that allows
\emph{linear programming} (LP) to help inference:
%
\jan{I wouldn't say that LP is the reason we picked this format. We
  rather picked it because it covers the most common cases.}
%
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom(\state)}
  q_{(x,y)} \cdot |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Here, we assume
that the program state is a simple map from program variables to integers
and we treat constants as program variables (that cannot be changed
by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
the sum.  We then develop an inference rule for each syntactic construct
that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
the previous section.

@@

\jan{I would add another subsection here to explain how we automate
  the analysis with LP solving.  Message: Efficient reduction to LP
  solving and LP solvers are great and super fast. Maybe mention
  'network problems'.}

\subsection{Compositionality in Practice}

With two concrete examples from open-source projects we demonstrate
that the compositionality of our method is indeed needed in the
analysis of common code.

\jan{I don't like to talk about ``the reader''. Often you can leave it
  out. It is fine to address the reader with ``you''. See BUGS in
  Writing.}

The following example is typical in cryptographic primitives
implementations.  These primitives work by block, so when data of
arbitrary length needs to be processed, it is munched in blocks and
the leftovers are stored in a buffer for future use when more data
is available.  This looping pattern is particularly well handled by
our tool because the LP solver is leveraged to generate a tight bound.
The optimization problem it is given encodes that constant bounds are
preferable to linear bounds.  Thus, if $N >= 8$, the bound given by
AAA is $\frac{N}{8} |[0,l]|$, but if $N<8$ it becomes $7\frac{8-N}{8}
+ \frac{N}{8}|[0,l]|$, because the cost for the second loop is better
counted as constant.
%
\jan{N is unbound above. The explanation must be much slower
  here. Explain what the code does and how the ticks account for the
  cost. Then: What is the resource bound?. Then: What do we derive?
  Btw: Why do you use tick(1) and not tick(M)?}
%
\begin{lstlisting}
for (; l >= 8; l -= 8)
    // process one block
    tick(N);
for (; l > 0; l--)
    // save leftovers
    tick(1);
\end{lstlisting}
%
It is worth taking some time to explain the constant contribution
to the resource bound when $N<8$.  Remark first that the cost of
the second loop is $|[0,l]|$, and after the first one, we still have
$\frac{N}{8}|[0,l]|$ potential available from the invariant.  So we have
to raise the potential of $|[0,l]|$ from $\frac{N}{8}$ to 1, that is,
we must pay $\frac{8-N}{8}|[0,l]|$.  But since we got out of the first
loop, we know that $l<8$, so it is sound to only pay $7\frac{8-N}{8}$
potential units instead.  This explains the somewhat intriguing value
of the constant term in the bound.  All this reasoning is done within
our tool automatically and simply relies on general rules presented
in further sections.

@@

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && nondet());
    do { h--; tick(1); } while (h > l && nondet());
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  However, all available
tools for C will derive a quadratic bound instead.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  But on the other hand, current
techniques will try to find one linear ranking-function for each loop
and combine them multiplicatively or additively, resulting in a loose
quadratic bound.
%
\jan{Did you test the last example with other tools? What are the
  bounds they get? State the bounds here.}
